---
title: "Bulk_rna_seq_GSE106305"
author: "Amina Hanaan"
date: "2025-08-22"
output: html_document
---
###Loading Packages
```{r}
library(tximport)
library(readr)
library(RColorBrewer)
library(pheatmap)
library(apeglm)
library(limma)
library(sva)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(tibble)

# install necessary libraries
#install.packages("BiocManager")
#BiocManager::install(version = "devel")
#BiocManager::install("ReactomePA")
```

###Setiing path 
```{r setup}
dir <- getwd()
work_dir <- normalizePath(file.path(dir, "/../"))
setwd(work_dir)
knitr::opts_knit$set(root.dir = work_dir)

# Create directory if it doesn't exist
dir.create(file.path(work_dir, "data"), recursive = TRUE)
dir.create(file.path(work_dir, "results"), recursive = TRUE)

source("scripts/visualisation.R")
source("scripts/degs_pathway_enrichment.R")


```

###Counts and metadata
Counts obtained from featureCounts are loaded and metadata created to be fed into Deseq
```{r}
#Loading counts after running featureCounts
raw_counts <- read.csv("data/GSE106305_counts_matrix.csv", header = TRUE, row.names = "Geneid", stringsAsFactors = FALSE)
raw_counts <- raw_counts[,sort(colnames(raw_counts))]
head(raw_counts)
colSums(raw_counts)

#setting up metadata
condition <- c(rep("LNCAP_Hypoxia", 2), rep("LNCAP_Normoxia", 2), rep("PC3_Hypoxia", 2), rep("PC3_Normoxia", 2))
print(condition)

my_colData <- as.data.frame(condition)
rownames(my_colData) <- colnames(raw_counts)

# Write metadata to CSV
write.csv(my_colData, file = "data/metadata.csv", row.names = FALSE)
write.csv(my_colData,file = "data/metadata.csv")

```

###Creating DESeq object
```{r}
install_if_missing("DESeq2")

dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = my_colData,
                              design = ~condition)
dds
head(counts(dds))
```
Genes with no gene expression across samples are checked. The table contains info in number of genes that have zero expression at different number of samples. 
```{r}
count_matrix <- counts(dds)
zero_counts_per_gene <- rowSums(count_matrix == 0)
count_matrix <- as.data.frame(count_matrix)
zero_summary <- table(zero_counts_per_gene)
print(zero_summary) #number of genes with diff. number of zero gene expression in samples
```

###Ensemble IDs to gene symbols using BioMart
Gene annotations (Ensembl ID, gene name, gene type) were downloaded via Ensembl BioMart <https://asia.ensembl.org/biomart/martview/03b16017eeb03b816d404a27eb7d9a55> and merged with the raw count matrix using Ensembl IDs.
```{r}
library(data.table)
library(dplyr)

annotation <- fread("data/GRCh38.p14_annotation.csv",stringsAsFactors = FALSE)
#renaming colnames
names(annotation)[names(annotation) %in% c("Gene stable ID", "Gene type", "Gene name")] <- c("Geneid", "Genebiotype", "Genesymbol")
annotation$Geneid <- sub("\\..*$","",annotation$Geneid)
head(annotation)
dim(annotation)
#matrix to dataframe

raw_counts <-read.csv("data/GSE106305_counts_matrix.csv",
                     header = TRUE,
                     stringsAsFactors = FALSE)
raw_counts$Geneid <- sub("\\..*$","",raw_counts$Geneid)
dim(raw_counts)
annotated_data <- left_join(raw_counts,annotation, by = "Geneid")
dim(annotated_data)
head(annotated_data)
write.csv(annotated_data, file = "data/gene_annotated_raw_counts.csv")
```

###Filtering
Filtered the annotated count matrix to retain only protein-coding and immune-related gene types, reducing noise and focusing on biologically relevant signals
```{r}
genetypes_to_keep <- c("protein_coding", "IG_J_gene", "IG_V_gene", "IG_C_gene", "IG_D_gene", "TR_D_gene", "TR_C_gene", "TR_V_gene", "TR_J_gene")

filtered_counts<- annotated_data[annotated_data$Genebiotype %in% genetypes_to_keep,]
dim(filtered_counts)
filtered_counts$Geneid <- sub("\\..*$", "", filtered_counts$Geneid)

output_file <- "data/genetypes_counts_matrix.csv"
fwrite(filtered_counts, file = output_file, sep = ",", row.names = FALSE)

```

Low-expression genes (zero counts in ≥7 samples) were removed to improve statistical power(2 replicates for each condition)
```{r}
zero_counts1 <- rowSums(filtered_counts[, 4:11] == 0)
zero_summary2 <- table(zero_counts1)
print(zero_summary2)

#keep genes if less than 7 samples have zero expression (since 2 replicates)
keep_genes <- zero_counts1 < 7
filtered_counts_nozero <- filtered_counts[keep_genes, ]
cat("Number of genes after filtering (zeros in <7 samples):", nrow(filtered_counts_nozero), "\n")

new_zero_counts <- rowSums(filtered_counts_nozero[, 4:11] == 0)
print(table(new_zero_counts)) #check if removed successfully 

output_file <- "data/genetype_nonzero_count_matrix.csv"
fwrite(filtered_counts_nozero, file = output_file, sep = ",",row.names= FALSE)
```

DeseqDataSet is filtered for only the genes left in the filtered counts. A table for genes that were removed from desired genetypes was created
```{r}
dds_filtered <- dds[rownames(dds) %in% filtered_counts_nozero$Geneid, ]
cat("Dimensions of filtered DESeqDataSet:", dim(dds_filtered), "\n")
head(dds_filtered)

#list of zero gene expression > 7 sample under diff. gene types
removed_genes <- filtered_counts[!keep_genes, ]
cat("Biotype distribution of removed genes:\n")
print(table(removed_genes$Genebiotype)) 
```
The distribution of biotypes left is visualised after filtering. 
```{r}

# Write metadata to CSV
write.csv(my_colData, file = "data/metadata.csv", row.names = FALSE)
install_if_missing("ggplot2")

genetype_counts <-filtered_counts_nozero %>%
  count(Genebiotype) %>%
  mutate(Proportion = n/sum(n),
         Percentage = Proportion * 100) %>%
  rename(Biotype = Genebiotype)

#Barplot
p <- ggplot(genetype_counts, aes(x = reorder(Biotype, -Proportion), y = Proportion, fill = Biotype)) +
  geom_bar(stat = "identity") +
  labs(title = "Proportion of Genes by Biotype",
       x = "Gene Biotypes",
       y = "Proportion") +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +  # Show proportions as percentages
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "none") +  # Remove legend if biotypes are clear
  scale_fill_brewer(palette = "Set2")  # Use distinct colors
p
output_plot <- "results/genebiotype_proportions.png"
ggsave(output_plot, bg = "white", plot = p, width = 8, height = 6, dpi = 300)
```

###PCA plot before DESeq
PCA clusters samples with similar gene expression before Deseq. The variable stabilising transformation is deals with genes that are highly expressed in small datasets.The plot helps us to visualise variance structure and potential batch effect
```{r}
#Stabilising variance across genes in sample
vsd <- vst(dds_filtered , blind = TRUE)

plot <- plot_PCA(vsd,"results/pca_before.png")
plot
```

### DESeq - for differential expressed genes
```{r}
dds <- DESeq(dds_filtered)
dds
saveRDS(dds, file = "normoxia_vs_hypoxia_deseq.rds")
normalized_counts <- counts(dds, normalized = T)
normalized_counts_df <- as.data.frame(normalized_counts)
write.csv(normalized_counts_df, file = "data/normalized_counts.csv", row.names = TRUE)
```

###PCA plot after DESeq
PCA to check how well the samples cluster by condition. This helps confirm that replicates behave consistently and that the major sources of variation are biologically meaningful.
```{r}

vsd <- vst(dds, blind = TRUE)  # blind=TRUE for exploratory PCA

plot_PCA(vsd,"results/pca_after.png")
plot
```

###Distance plot 
- calculates Euclidean distance based on the expression values of genes in order to check how similar the expression profiles are across conditions. This helps spot outliers and confirms that replicates cluster as expected
```{r}
plot<-plotDists(vsd)
plot
```

###Variable genes HeatMap
The genes that drives the clustering can be visualised through Heatmap. Here top 40 variable genes are taken into account.
```{r}
plot<-variable_gene_heatmap(vsd, num_genes = 40, annotation = annotation)
plot

```
###Density plots
Plotted density curves for raw and VST-transformed counts across all samples to check how well variance was stabilized. This helps confirm that expression distributions are more comparable post-transformation.
```{r}
raw_counts <- assay(dds)
vst_counts <- assay(vsd)

png("results/density_plots_raw_vst.png",
    width = 4000, height = 4000, res = 300)  # Adjust width, height (pixels), and resolution (dpi)

par(mfrow = c(4, 4), mar = c(3, 3, 2, 1))  # mar adjusts margins (bottom, left, top, right)

for (i in 1:8) {
  # Raw counts density
  plot(density(raw_counts[, i]),
       main = paste("Raw - Sample", colnames(raw_counts)[i]),
       xlab = "Expression",
       col = "red",
       lwd = 2,
       ylim = c(0, max(sapply(1:8, function(j) max(density(raw_counts[, j])$y, na.rm = TRUE))))  # Uniform y-axis
       )
  
  # VST counts density (next panel)
  plot(density(vst_counts[, i]),
       main = paste("VST - Sample", colnames(vst_counts)[i]),
       xlab = "Expression",
       col = "blue",
       lwd = 2,
       ylim = c(0, max(sapply(1:8, function(j) max(density(vst_counts[, j])$y, na.rm = TRUE))))  # Uniform y-axis
       )
}
dev.off()
```
```{r}
gene_plot<-gene_plot_counts(dds, "IGFBP1")
gene_plot
```


```{r}
sink("preprocess_session_info.txt")
sessionInfo()
sink()
```

*THE END*