---
title: "LNCAP - Normoxia vs Hypoxia"
author: "Amina Hanaan"
date: "2025-11-14"
output: html_document
---
###Setiing path 
```{r setup}
dir <- getwd()
work_dir <- normalizePath(file.path(dir, "/../"))
setwd(work_dir)
knitr::opts_knit$set(root.dir = work_dir)

# Create directory if it doesn't exist
dir.create(file.path(work_dir, "data"), recursive = TRUE)
dir.create(file.path(work_dir, "results"), recursive = TRUE)

source("scripts/degs_pathway_enrichment.R")
```

#LNCAP - Hypoxia VS Normoxia
The LNCAP sampled were filtered from dataset, to compare hypoxia vs normoxia. Normoxia is set as reference and DESeq2 was ran. The genes are ordered based on adjusted p-values
```{r}
cell_type <- "lncap"
loaded_dds <- readRDS("normoxia_vs_hypoxia_deseq.rds")
# Filter the DESeq2 dataset (dds) to keep only LNCAP samples
cell <- filter_celltype(dds,condition)
dds_cell <- cell$dds
res_cell <- cell$results
rescellOrdered<-cell$ordered_results
```


```{r}
plotMA(rescellOrdered,ylim=c(-2,2))
```
###Volcano plot - gene regulation
The gene regulation under hypoxia condition in LNCAP cell lines, based on which volcano plot is created. Genes were categorized as upregulated, downregulated, or not significant based on fold change and adjusted p-value.
```{r}
plot <- gene_regulation_plot(rescellOrdered)
plot$gene_plot
```

###Gene Set Enrichment Analysis (GSEA)
GSEA was performed using ReactomePA on DESeq2-derived log2 fold changes from LNCaP cells under hypoxia versus normoxia. Ensembl IDs were mapped to Entrez IDs, and genes were ranked by fold change. Enriched pathways were identified based on normalized enrichment scores (NES) and adjusted p-values.
```{r}
head(res_cell)

res <- convert_to_entrez(res_cell)
res_mapped<-res$res_mapped
ngenes <- res$ngenes

#GSEA
enp_gsea <- run_gsea(ngenes)

head(enp_gsea@result)
```
GSEA results are sorted by adjusted p-value and normalized enrichment score (NES) to highlight the most responsive pathways. Here entire expression profile (including non-significant genes) is considered
```{r}
top20<- top_20_pathways(enp_gsea)
```

###Pathway enrichment based on overall expression profile
Plotted the top Reactome pathways from GSEA using NES, FDR, and gene set size to show which biological processes are most enriched
```{r}
pathway_enrich_overall(top20)
```

###Pathway enrichment by significant DEGs
Pathway enrichment using ReactomePA to see which biological processes are impacted in the significant DEGs
```{r}
pathway_enrich_sigDEGS(res_mapped, organism = "human")
```

###GSEA of Hallmark Programs 
in order to understand broad biological responses in LNCAP cells under hypoxia

Ranked Gene List for GSEA using log2fold change
```{r}
res_cell <- read.csv(paste0("D:/BI_prj/bulkrnaseq_proj/data/DEGs_",cell_type,".csv"), row.names = 1)
head(res_cell)
colnames(res_cell)

rank_df <- ranked_gene_list(res_cell)

```


Downloaded Hallmark gene sets from MSigDB <https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp#H> and prepped the ranked list for GSEA
```{r}
library(fgsea)
# Load Hallmark gene sets from GMT file
hallmark_pathway <- gmtPathways("D:/BI_prj/bulkrnaseq_proj/normoxia_vs_hypoxia/data/h.all.v2025.1.Hs.symbols.gmt")
head(names(hallmark_pathway))
head(hallmark_pathway$HALLMARK_HYPOXIA, 20)

```

###Validate Ranked Gene list
```{r}
str(rank_df)
ranked_list<- prepare_ranked_list(rank_df)


print(head(ranked_list))
```

Run GSEA using Hallmark gene sets and ranked LNCaP gene list
```{r}
cell_fgsea<-run_fgsea(hallmark_pathway,ranked_list)
fgsea_results<-cell_fgsea$results  
dim(fgsea_results)

fgsea_results_ordered<-cell_fgsea$ordered_results

```

###Hallmark GSEA – Hypoxia Response
The plot to show which Hallmark programs are enriched in LNCaP cells under hypoxia. Pathways are sorted by NES, and bars are colored by significance (padj < 0.05).

```{r}
hallmark_enrich<- waterfall_plot(fgsea_results, paste0("Hallmark pathways altered by hypoxia in ",toupper(cell_type)," cells"))

```

```{r}
sink("lncap_session_info.txt")
sessionInfo()
sink()
```
